---
title: "AN588-Week-2"
author: "Gentry Miller"
format: html
editor: visual
date: "23 Jan, 2025"
editor_options: 
  chunk_output_type: console
---

```{r}
install.packages("stringr", repos = "http://cran.us.r-project.org")
library("stringr")
```

# Module 4

## Challenge 1

```{r}
v1 <- c("Stately,", "plump", "Buck", "Mulligan", "came", "from", "the", "stairhead,", "bearing", "a", "bowl", "of", "lather", "on", "which", "a", "mirror", "and", "a", "razor", "lay", "crossed")
#v1
```

```{r}
# I tried to use str_split() and str_split_i() but it was only pulling characters instead of the whole sub-string. When I set the boundary to word, it pulled NA.
v2 <- v1[seq(0, length(v1), 3)]
v2
gsub("[[:punct:]]", "", v1)
```

## Challenge 2

```{r}
m <- matrix(data = 1:80, nrow = 8, ncol = 10, byrow = FALSE)
```

```{r}
x <- matrix(data = c(m[,2], m[,3], m[,6]), nrow = 8, ncol = 3, byrow = FALSE)
```

```{r}
x <- matrix(data = m[6:8,], nrow = 3, ncol = 10, byrow = FALSE) 
```

```{r}
x <- matrix(data = m[2:6,2:9], nrow = 5, ncol = 8)
```

## Challenge 3

```{r}
a <- array(400:1, dim = c(5, 5, 4, 4))
```

```{r}
a[1, 1, 1, 2]
a[2, 3, 2,]
a[1:5, 1:5, 3, 3]
```

### 3.1

`a[1, 1, 1, 2]` returns 300

### 3.2

`a[2, 3, 2, ]` returns a vector containing 364, 264, 164, and 64

### 3.3

`a[1:5, 1:5, 3, 3]` returns a 5x5 matrix of the first 5 values for the first and second dimensions of a at the third value of the third and fourth dimensions.

## Overwriting

```{r}
# Replaces a single val at m[7,1]
m[7, 1] <- 564
# Replaces the 8th column of m with all 2's
m[, 8] <- 2
# Replaces the values in rows 2-5 and columns 4-8 with 1
m[2:5, 4:8] <- 1
# Replaces the vals in row 2 col 4-8 with 20, row 3 cols 4-8 with 19, etc.
m[2:5, 4:8] <- c(20, 19, 18, 17)
# Replaces rows 2-5 and cols 4-8 with a new matrix of decreasing ints from 20-1 by row
m[2:5, 4:8] <- matrix(data = c(20:1), nrow = 4, ncol = 5, byrow = TRUE)
# replaces the vals in col 8 with alternating chars a and b and changes the data type of every value in the matrix to character
m[, 8] <- c("a", "b")
```

## Lists and Data Frames

```{r}
s <- c("this", "is", "a", "vector", "of", "strings")
m <- matrix(data=1:40, nrow=5, ncol=8)
b <- FALSE
l <- list(s, m, b)
```

```{r}
l[[2]]
```

```{r}
l[[2]][2, 6]
```

```{r}
l[c(1,3)]
```

```{r}
str(l)
names(l) <- c("string", "matrix", "logical")
names(l)
l$string
l$matrix[3, 5]
```

## Challenge 4

### Phylogeny

```{r}
# Anthropoidea
Platyrrhini <- c("Cebidae", "Atelidae", "Pitheciidae")
Catarrhini <- c("Cercopithecidae", "Hylobatidae", "Hominidae")
Anthropoidea <- list(Platyrrhini, Catarrhini)
names(Anthropoidea) <- list("Platyrrhini", "Catarrhini")

# Tarsioidea
Tarsioidea <- c("Tarsiidae")

# Haplorhini
Haplorhini <- list(Anthropoidea, Tarsioidea)
names(Haplorhini) <- c("Anthropoidea", "Tarsioidea")

# Lorisoidea
Lorisoidea <- c("Lorisidae", "Galagidae")

# Lemuroidea
Lemuroidea <- c("Cheirogaleidae", "Lepilemuridae", "Indriidae", "Lemuridae", "Daubentoniidae")

# Strepsirhini
Strepsirhini <- list(Lorisoidea, Lemuroidea)
names(Strepsirhini) <- c("Lorisoidea", "Lemuroidea")

# Primates
Primates <- list(Haplorhini, Strepsirhini)
names(Primates) <- c("Haplorhini", "Strepsirhini")

Primates
```

### Data Frames

```{r}
df <- read.csv("C:/Users/mille/Documents/School/BU Undergrad/Fifth Year/Spring25/AN588/AN588-Modules/random-people.csv", sep = ",", header = TRUE, stringsAsFactors = FALSE)
# only print select columns
df[, c(1, 3, 4, 11, 12)]
```

```{r}
str(df)
df[,3]
str(df[,3])
df$name.last
df[3]
str(df[3])
df["name.last"]
str(df["name.last"])
```

```{r}
df <- cbind(df, id = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16,
    17, 18, 19, 20))
df <- cbind(df, school = c("BU", "BU", "Harvard", "Harvard", "BU", "Boston College",
    "MIT", "BU", "BU", "UMASS Boston", "Harvard", "BU", "Boston College", "BU",
    "Harvard", "MIT", "Harvard", "BU", "UMASS Boston", "Harvard"))
df
```

### Filtering

```{r}
new_df <- df[df$school == "BU",]
new_df
```

```{r}
new_df <- df[df$school == "BU", c("name.last", "name.first", "school")]
new_df
```

## Challenge 5

```{r}
# Create a vector to inform the matrix
md = c(3, 0, 1, 23, 1, 2, 33, 1, 1, 42, 0, 1, 41, 0, 2)
# Organize the data from md vector into matrix
m3 <- matrix(data = md, nrow = 5, ncol = 3, byrow = TRUE)

str(m3)
```

```{r}
df <- data.frame(m3)
df$X2 <- as.logical(df$X2)
df$X3 <- as.factor(df$X3)
str(df)
```

**\*Jon's Comments**

Challenge 1:

-   You may need to first split the sentence into each words. Try \*\*function \<- v1\[seq(1, length(v1), by = 3)\]

Challenge 2-3

-   Looks pretty good

Challenge 4

-   There is a error with the code for the Haplorhini. I think it just requires a change in the names function between Haplothini and Tarsioidea.

Challenge 5

-   Looks pretty good there.
